---
title: "**POLI706: Advanced Methods of Political Analysis**"
subtitle: |
          | 
          | Problem set 5
          |
documentclass: article
format: 
  pdf:
    citation_package: natbib
    keep_tex: true
    fig_caption: true
    latex_engine: xelatex
    code-block-bg: light
    code-block-border-left: "#31BAE9"
    linkcolor: highlight
    urlcolor: highlight
    highlight-style: atom-one
#bibliography: W2.bib
header-includes:
  -  \usepackage{hyperref}
  -  \usepackage{mathpazo}
  -  \usepackage{bookmark}
  -  \definecolor{light}{HTML}{EEEEEE}
  -  \definecolor{highlight}{HTML}{9A2A2A}
  -  \definecolor{dark}{HTML}{330033}
  -  \usepackage{setspace}\singlespacing
  -  \usepackage{fvextra}
  -  \setmainfont[ItalicFont = {TeXGyrePagella-Italic}]{TeX Gyre Pagella}
  -  \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
  #-  \setmonofont{Fira Code}[Contextuals=Alternate]
  -  \makeatletter
  -  \renewcommand*\verbatim@nolig@list{}
  -  \makeatother
# You have to install fira font and tex gyre pagella
biblio-style: apsr
geometry: margin=1in
#fontfamily: mathpazo
fontsize: 11pt
link-citation: true
indent: true
---

```{r}
#| include: false
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, error = FALSE, 
                      message = FALSE, warning = FALSE)
# Load packages to import data
library(ezpickr);library(knitr);library(patchwork);library(tidyverse)
```

## Exercise 1

For this exercise we will use the `cats` dataset from the `MASS` package. You should use `?cats` to learn about the background of this dataset.

```{r}
library(MASS)
data(cats)
```


a. Fit the following simple linear regression model in `R`. Use heart weight as the response and body weight as the predictor. 

$$ Y_i = \beta_0 + \beta_1 x_i + \epsilon_i$$


Store the results in a variable called `cat_model`. Use a $t$ test to test the significance of the regression. Report the following:

b. The null and alternative hypotheses

c. The value of the test statistic

d. The p-value of the test

e. A statistical decision at $\alpha = 0.05$

f. A conclusion in the context of the problem

```{r}
cat_model = lm(Hwt ~ Bwt, data = cats);
summary(cat_model)
```

## Exercise 2

What are $\beta_0$, $\beta_1$, $\hat{\beta_0}$, $\hat{\beta_1}$, $x$, $x_i$, $u$, $\hat{u_i}$ in output of simple linear regression from the Exercise 1? 

## Exercise 3
Explain under which circumstances you will get a large slope coefficient estimates when you run a regression model.