---
title: "**POLI706: Advanced Methods of Political Analysis**"
subtitle: |
          | 
          | Problem set 9
          |
documentclass: article
format: 
  pdf:
    citation_package: natbib
    keep_tex: true
    fig_caption: true
    latex_engine: xelatex
    code-block-bg: light
    code-block-border-left: "#31BAE9"
    linkcolor: highlight
    urlcolor: highlight
    highlight-style: atom-one
bibliography: ../references/references.bib
header-includes:
  -  \usepackage{hyperref}
  -  \usepackage{mathpazo}
  -  \usepackage{bookmark}
  -  \definecolor{light}{HTML}{EEEEEE}
  -  \definecolor{highlight}{HTML}{9A2A2A}
  -  \definecolor{dark}{HTML}{330033}
  -  \usepackage{setspace}\singlespacing
  -  \usepackage{fvextra}
  -  \setmainfont[ItalicFont = {TeXGyrePagella-Italic}]{TeX Gyre Pagella}
  -  \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
  #-  \setmonofont{Fira Code}[Contextuals=Alternate]
  -  \makeatletter
  -  \renewcommand*\verbatim@nolig@list{}
  -  \makeatother
# You have to install fira font and tex gyre pagella
biblio-style: apsr
geometry: margin=1in
#fontfamily: mathpazo
fontsize: 11pt
link-citation: true
indent: true
---

```{r}
#| include: false
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, error = FALSE, 
                      message = FALSE, warning = FALSE)
# Load packages to import data
library(ezpickr);library(knitr);library(patchwork);library(tidyverse)
```

## Exercise 1

Read @brambor2006, and take notes. Lots of them.

## Exercise 2

Read @king2000, and take notes. Lots of them.

## Exercise 3

Some scholars have posited the "More Murder in the Middle" (MMM) effect, suggesting that as democracy develops, the level of human rights repression initially increases, but after surpassing a certain threshold, it begins to decrease. To verify the MMM hypothesis, perform a regression analysis and create a graph that includes the confidence intervals to illustrate the estimated levels of human rights based on different levels of democracy. Clearly identify the turning point in your analysis.

1.  Run the following R codes and load necessary packages to use.

2.  Load sample data for the assignment: `data_assignment9.csv`

3.  Use base R functions to answer the questions.

| Variables       | Description                                            | Source                       |
|------------------|-----------------------------------|-------------------|
| `country_name`  | Name of country                                        | V-Dem                        |
| `COWcode`       | Country code of Correlates of War                      | V-Dem                        |
| `year`          | Year                                                   | V-Dem                        |
| `hrs`           | Human Rights Scores                                    | See @fariss2014              |
| `v2x_polyarchy` | Electoral Democracy Index                              | V-Dem                        |
| `e_gdppc`       | Estimated GDP per capita                               | V-Dem                        |
| `e_pop`         | Estimated population                                   | V-Dem                        |
| `wdi_trade`     | Trade openness measured by (imports + exports)/GDP (%) | World Development Indicators |
| `ucdp_type3`    | Number of interstate armed conflicts                   | UCDP Dyadic Dataset          |

: Variables in the sample data

a.  Take the natural logarithm of GDP per capita and population, and store the new variables in your sample. Then, make a binary variable of civil war experience indicating if a country experiences internal conflict once, code 1 otherwise code 0.

```{r}
#| echo: false
#| eval: true
#| results: hide
data_assignment9 <- read_csv("data_assignment9.csv")
data_assignment9 |> 
  mutate(lngdppc = log(e_gdppc),
         lnpop = log(e_pop),
         civilwar = if_else(is.na(ucdp_type3) | ucdp_type3 == 0L, 0L, 1L)) ->
  assignment9

```

b.  Estimate a model to test the MMM effect using the sample data, which takes the form of $Y = \beta_0 + \beta_1 X + \beta_2X^2 + \beta_3 \mathrm{Controls} + \epsilon$. . Provide the results in a table with appropriate labels and format. Explain how the model can test the MMM effect.

```{r}
#| echo: false
#| eval: true
#| results: hide
lm(hrs ~ v2x_polyarchy + I(v2x_polyarchy^2) +
     lngdppc + lnpop + civilwar, data =  assignment9) -> mmm

mmm |> texreg::screenreg()

```

b.  Re-read @brambor2006 and @king2000. From the Question 2a, visualize the results to ilustrate the marginal effect of democracy on human rights. Clearly identify the turning point in your analysis.

c.  Re-read @brambor2006 and @king2000. From the Question 2a, visualize the results to llustrate the estimated levels of human rights based on different levels of democracy. Clearly identify the turning point in your analysis. **Hint!** You can estimate the levels of human rights based on different levels of democracy holding other covariates constant.

## Exercise 4

According to @hainmueller2019, what should you do next?

## References
